Statistical Analysis

### 2025-05-03 code/analysis-code/statistical-analysis.qmd
Part 6

---
title: 
author: 
date: 
output: html_document
---
# loading in packages
```{r}
library(tidyverse)
library(here) 
library(dplyr)
library(skimr)
library(ggplot2)
library(tidymodels)
library(lubridate)
library(ranger)
library(parsnip)
library(poissonreg)
library(randomForest)
```
#Load in the data
```{r}
#Path to data. 
data_location <- here::here("data","processed-data","processeddata.rds")
#load data
mydata <- readRDS(data_location)
```

#Here I will perform some basic statistical analysis.

#Lets start with a basic glm of time and food and time and clinical. We will also do some visualizations using a histrogram

#To do this I have to convert my date variable into a numeric variable.

```{r}
# Load necessary package
library(lubridate)

"# Sample data
mydata <- data.frame(
  date `Create date` = as.Date,
  outcome = clinical  # Yes = 1, No = 0
)"

# Recode Yes = 1, No = 0
mydata$clinical_numeric <- ifelse(mydata$clinical == "Yes", 1, 0)

# Convert date to numeric (days since first observation)
mydata$days_since_start <- as.numeric(difftime(mydata$`Create date`, min(mydata$`Create date`), units = "days"))

# Logistic regression
model <- glm(clinical_numeric ~ days_since_start, data = mydata, family = binomial)

# Model summary
summary(model)

@@

# temporal analysis of clinical isolates. Remember that every clinical isolate is like a human case of Listeriosis.
#For this analysis I will make some changes to my data by separating the time data into seasons.Winter is considered December, January and February; spring is March through May; summer is June through August; and fall or autumn is September through November.

```{r}
mydata <- mydata %>%
  drop_na()%>%
  mutate(
    `Create date` = as.Date(`Create date`),
    year = year(`Create date`),
    day_of_year= yday(`Create date`),
    week = week(`Create date`),
    month = month(`Create date`),
    season = case_when(
      month %in% c(12, 1, 2)  ~ "Winter",
      month %in% c(3, 4, 5)   ~ "Spring",
      month %in% c(6, 7, 8)   ~ "Summer",
      month %in% c(9, 10, 11) ~ "Fall"
    ),
    season = factor(season, levels = c("Winter", "Spring", "Summer", "Fall")),
    `Isolation source` = factor(`Isolation source`)
  )
```

#Feature selection
```{r}
model1_data <- mydata %>%
  select(`Isolation type`,year,month,day_of_year,week,season)
```

#Split the data
```{r}
set.seed(1234)
data_split <- initial_split(model1_data, prop = 0.8, strata = `Isolation type`)
train_data <-training(data_split)
test_data <-testing(data_split)
```
#Create recipe
```{r}
rf_recipe <- recipe(`Isolation type` ~ month + year + day_of_year + week + season, data = train_data) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_zv(all_predictors()) %>%
  step_normalize(all_numeric_predictors())
```

#Define the model
```{r}
rf_mod <-rand_forest(trees = 150, min_n =5) %>%
  set_mode("classification") %>%
  set_engine("ranger",importance="impurity")
```
Create workflow
```{r}
workflow1 <- workflow() %>%
  add_model(rf_mod)%>%
  add_recipe(rf_recipe)
```
#fit the random forest model
```{r}
fit1 <- workflow1 %>%
  fit(data = train_data)
```
#Make predictions for evaluation
```{r}
predictions_rf <-fit1 %>%
  predict(new_data = test_data) %>%
  bind_cols(test_data)

```
```{r}
#set isolation source as a factor
predictions_rf <- predictions_rf %>%
  mutate(`Isolation type` = as.factor(`Isolation type`))

@@

#Run metrics to evaluate the model
```{r}
# Create a metric set for classification
classification_metrics <- metric_set(accuracy, precision, recall, f_meas)
```

```{r}
#  use it on predictions
metrics_rf <- predictions_rf %>%
  classification_metrics(truth = `Isolation type`, estimate = .pred_class)

@@

```{r}
metrics_rf <- predictions_rf %>%
  metrics(truth = `Isolation type`, estimate = .pred_class)
print(metrics_rf)
```
#Plot the results of this model

```{r}
library(vip)  # for variable importance plot

# Extract the fitted model from the workflow
rf_fit <- extract_fit_parsnip(fit1)

# Plot variable importance
vip(rf_fit$fit)

@@

#Poisson modeling using month

```{r}
# Create monthly clinical counts
monthly_counts <- mydata %>%
  filter(clinical == "Yes") %>%
  mutate(month = floor_date(`Create date`, "month")) %>%
  group_by(month) %>%
  summarize(clinical_count = n()) %>%
  ungroup()
#filling in gaps with 0
monthly_counts <- monthly_counts %>%
  complete(month = seq.Date(min(month), max(month), by = "month"),
           fill = list(clinical_count = 0))
#Feature engineering, extracting the month from the data
monthly_counts <- monthly_counts %>%
  mutate(month_name = factor(month(month, label = TRUE), ordered = FALSE))

@@

```{r}
library(tidymodels)
library(parsnip)
#Splitting the data 
set.seed(123)
data_split2 <- initial_time_split(monthly_counts, prop = 0.8)
train_data2 <- training(data_split2)
test_data2  <- testing(data_split2)

```
```{r}
#create a recipe 
poisson_recipe <- recipe(clinical_count ~ month_name, data = train_data2) %>%
  step_dummy(all_nominal_predictors())

```
```{r}
#Select the model
poisson_model <- 
  poisson_reg(
    mode = "regression",
    penalty = NULL,
    mixture = NULL,
    engine ="glm"
  )

```
```{r}
#Create a workflow 
library(poissonreg)
poisson_wf <-workflow() %>%
  add_model(poisson_model) %>%
  add_recipe(poisson_recipe)
```
```{r}
#fit the model
poisson_fit <- poisson_wf %>%
  fit(data = train_data2)
```
```{r}
#make predictions
predictions <- predict(poisson_fit, new_data = train_data2) %>%
  bind_cols(train_data2)
metrics(predictions, truth = clinical_count, estimate =.pred)
```
#Visualize results
```{r}
poisson_seasonal <- ggplot(predictions, aes(x = month)) +
  geom_line(aes(y = clinical_count), color = "black", linetype = "dashed") +
  geom_line(aes(y = .pred), color = "blue") +
  labs(title = "Predicted vs Actual Monthly Clinical Isolate Counts",
       y = "Clinical Count")
print(poisson_seasonal)
ggsave("/Users/alexisgonzalez/Desktop/MADA/Gonzalez-MADA-project/results/figures/poisson_seasonal.png", plot = poisson_seasonal, width = 8, height = 6, dpi = 300)
```
#Predict future isolate counts based on seasonality
```{r}
last_month <-max(monthly_counts$month)
```

```{r}
future_months <- tibble(
  month = seq.Date(from = last_month %m+% months(1),by = "month", length.out = 12)
)%>%
  mutate(month_name = factor(month(month, label = TRUE), ordered= FALSE))
```

```{r}
#make predictions
future_predictions <- predict(poisson_fit, new_data = future_months)%>%
  bind_cols(future_months)
```
```{r}
#visualize
# Combine historical + future for plotting
combined_months<- bind_rows(
  monthly_counts %>% select(month, clinical_count) %>% mutate(source = "Actual"),
  future_predictions %>% rename(clinical_count = .pred) %>% mutate(source = "Forecast")
)

# Plot
seasonalpreds <- ggplot(combined_data, aes(x = month, y = clinical_count, color = source)) +
  geom_line() +
  labs(title = "Actual and Forecasted Clinical Isolate Counts",
       y = "Clinical Count",
       x = "Month") +
  scale_color_manual(values = c("Actual" = "black", "Forecast" = "blue")) +
  theme_minimal()
print(seasonalpreds)
ggsave("/Users/alexisgonzalez/Desktop/MADA/Gonzalez-MADA-project/results/figures/seasonalpreds.png", plot = seasonalpreds, width = 8, height = 6, dpi = 300)
```

```{r}

# 1. Get the last month from your historical data
last_month <- max(monthly_counts$month)

# 2. Create the next 12 months
future_months <- tibble(
  month = seq.Date(from = last_month %m+% months(1), by = "month", length.out = 12)
) %>%
  mutate(month_name = factor(month(month, label = TRUE), ordered = FALSE))

# 3. Make predictions using your Poisson model (or other model)
future_predictions <- predict(poisson_fit, new_data = future_months) %>%
  bind_cols(future_months)

# 4. Combine historical and forecast data
combined_months <- bind_rows(
  monthly_counts %>% 
    select(month, clinical_count) %>% 
    mutate(source = "Actual"),
  future_predictions %>% 
    rename(clinical_count = .pred) %>% 
    mutate(source = "Forecast")
)

# 5. Plot
seasonalpreds <- ggplot(combined_months, aes(x = month, y = clinical_count, color = source)) +
  geom_line(size = 1.2) +
  labs(
    title = "Actual and Forecasted Clinical Isolate Counts",
    y = "Clinical Count",
    x = "Month"
  ) +
  scale_color_manual(values = c("Actual" = "black", "Forecast" = "blue")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# 6. Show plot
print(seasonalpreds)

# 7. Save to file
ggsave(
  "/Users/alexisgonzalez/Desktop/MADA/Gonzalez-MADA-project/results/figures/seasonalpreds.png", 
  plot = seasonalpreds, width = 8, height = 6)

```{r}
library(mgcv)
gam_model <- gam(clinical_count ~ s(day_of_year), family = poisson(), data = daily_counts)
plot(gam_model)
```

```{r}
# Load necessary library
library(mgcv)
library(gratia) # for tidy ggplot GAM visualizations

# Example: Create ggplot-style plot from your GAM model
gam_plot <- draw(gam_model)  # `gam_model` is your fitted model

# Now save it
ggsave("/Users/alexisgonzalez/Desktop/MADA/Gonzalez-MADA-project/results/figures/gam_model.png",
       plot = gam_plot, width = 8, height = 6, dpi = 300)

```{r}

#  Split the data into training and testing sets
set.seed(123)
data_split <- initial_split(monthly_counts, prop = 0.8)
train_data <- training(data_split)
test_data <- testing(data_split)

#  Define the recipe (clinical_count predicted by food_count)
poisson_recipe <- recipe(clinical_count ~ food_count, data = train_data)

# Define the Poisson regression model
poisson_model <- poisson_reg(mode = "regression", engine = "glm")

#  Create a workflow 
poisson_wf <- workflow() %>%
  add_recipe(poisson_recipe) %>%
  add_model(poisson_model)

#  Fit the model
poisson_fit <- poisson_wf %>%
  fit(data = train_data)

# Make predictions on the training data
predictions <- predict(poisson_fit, new_data = train_data) %>%
  bind_cols(train_data)

#  Evaluate the model 
isolate_metrics <- metrics(predictions, truth = clinical_count, estimate = .pred)
print(isolate_metrics)

# Visualize the results
ggplot(predictions, aes(x = month)) +
  geom_col(aes(y = clinical_count), fill = "lightgray", alpha = 0.8) +
  geom_line(aes(y = .pred), color = "blue", size = 1) +
  labs(
    title = "Poisson Regression: Predicting Clinical Isolates from Food Isolates",
    y = "Clinical Isolate Count",
    x = "Month"
  ) +
  theme_minimal()
```
#Chart showing likelihood
```{r}

#Now I want to do one more advanced statistical analysis. This analysis will utilize the counts of clinical and counts of food isolates each year and see if the amount of clinical isolates can be predicted by the amount of food isolates. This purpose of this analysis is to see if federal reguulations are doing a good job of preventing illness.

#put my food and clinical variable into counts and group by year

```{r}
monthly_counts2 <- mydata %>%
  mutate(month = floor_date(`Create date`, "month")) %>%
  count(month, isolation_source) %>%
  tidyr::pivot_wider(names_from = isolation_source, values_from = n, values_fill = 0)
```

```{r}
set.seed(123)  # for reproducibility

rf_model <- randomForest(
  clinical ~ food + environmental,
  data = monthly_counts2,
  ntree = 500,          # number of trees
  importance = TRUE     # track variable importance
)
print(rf_model)

```{r}
#model_predict_clinical <- lm(clinical ~ food + environmental, data = monthly_counts2)
#summary(model)

```{r}

`
#Visualzing the future
```{r}
# Suppose you created this:
future_data <- data.frame(
  month = as.Date(c("2025-02-01","2025-03-01","2025-04-01","2025-05-01", "2025-06-01", "2025-07-01","2025-08-01","2025-09-01","2025-10-01","2025-11-01","2025-12-01")),
  food = c(300,220, 180, 250,140,122,116,235,96,175,225),
  environmental = c(350,150, 170, 200,220,140,267,178,155,200,280)
)
```

```{r}
# Predict clinical isolates
future_data$Predicted_Clinical <- predict(rf_model, newdata = future_data)
```

```{r}
# Predict clinical isolates for future months
future_data$Predicted_Clinical <- predict(rf_model, newdata = future_data)

# Combine with past data (no predicted clinicals in the past, so we set it to NA)
combined_plot_data <- monthly_counts2 %>%
  select(month, clinical) %>%
  mutate(Predicted_Clinical = NA,
         Source = "Historical") %>%
  bind_rows(
    future_data %>%
      transmute(month, clinical = NA, Predicted_Clinical, Source = "Forecast")
  )

# Filter combined_plot_data for the last 3 years
recent_data <- combined_plot_data %>%
  filter(month >= (max(month) %m-% years(3)))

>>>>

---
title: "Statistical Analysis"
author: "Alexis Gonzalez"
date: 
output: html_document
---

# loading in packages

```{r}
library(tidyverse)
library(here) 
library(dplyr)
library(skimr)
library(ggplot2)
library(tidymodels)
library(lubridate)
library(ranger)
library(parsnip)
library(poissonreg)
library(MASS)
```

#Load in the data

```{r}
#Path to data. 
data_location <- here::here("data","processed-data","processeddata.rds")
#load data
mydata <- readRDS(data_location)
```

# temporal analysis of clinical isolates. Remember that every clinical isolate is like a human case of Listeriosis.

#For this analysis I will make some changes to my data by separating the time data into seasons.Winter is considered December, January and February; spring is March through May; summer is June through August; and fall or autumn is September through November.

#POISSON MODELING
```{r}
# Create monthly counts for clinical, food, and environmental
monthly_clinical <- mydata %>% 
  filter(clinical == "Yes") %>% 
  mutate(month = floor_date(`Create date`, "month")) %>% 
  group_by(month) %>% 
  summarize(clinical_count = n()) %>% 
  ungroup()

monthly_food <- mydata %>% 
  filter(food == "Yes") %>% 
  mutate(month = floor_date(`Create date`, "month")) %>% 
  group_by(month) %>% 
  summarize(food_count = n()) %>% 
  ungroup()

monthly_environmental <- mydata %>% 
  filter(isolation_source == "environmental") %>% 
  mutate(month = floor_date(`Create date`, "month")) %>% 
  group_by(month) %>% 
  summarize(environmental_count = n()) %>% 
  ungroup()

# Combine all counts into one dataset
monthly_counts_combined <- monthly_clinical %>%
  full_join(monthly_food, by = "month") %>%
  full_join(monthly_environmental, by = "month")

# View the combined dataset
head(monthly_counts_combined)

@@

#TEMPORAL x CLINCAL

```{r}
# Filling in gaps with 0 for all counts
monthly_counts_combined <- monthly_counts_combined %>%
  complete(month = seq.Date(min(month), max(month), by = "month"),
           fill = list(clinical_count = 0, food_count = 0, environmental_count = 0))

# Feature engineering: extracting the month from the data
monthly_counts_combined <- monthly_counts_combined %>%
  mutate(month_name = factor(month(month, label = TRUE), ordered = FALSE))

# Splitting the data
set.seed(123)
data_split2 <- initial_time_split(monthly_counts_combined, prop = 0.8)
train_data2 <- training(data_split2)
test_data2  <- testing(data_split2)

# Create a recipe for clinical_count
poisson_recipe1 <- recipe(clinical_count ~ month_name, data = train_data2) %>%
  step_dummy(all_nominal_predictors())

# Select Poisson model
poisson_model <- poisson_reg(
  mode = "regression",
  penalty = NULL,
  mixture = NULL,
  engine = "glm"
)

# Create a workflow
library(poissonreg)
poisson_wf1 <- workflow() %>%
  add_model(poisson_model) %>%
  add_recipe(poisson_recipe1)

# Fit the model
poisson_fit1 <- poisson_wf1 %>% fit(data = train_data2)

# Make predictions
predictions1 <- predict(poisson_fit1, new_data = train_data2) %>%
  bind_cols(train_data2)

# Evaluate performance
library(yardstick)
metric_table <- metrics(predictions1, truth = clinical_count, estimate = .pred)
save_table_location <- here::here("results","tables","metric_table.rds")
saveRDS(metric_table, file = save_table_location)
# Predict future clinical isolate counts
last_month <- max(monthly_counts_combined$month)

future_months1 <- tibble(
  month = seq.Date(from = last_month %m+% months(1), by = "month", length.out = 12)
) %>%
  mutate(month_name = factor(month(month, label = TRUE), ordered = FALSE))

# Future predictions
future_predictions1 <- predict(poisson_fit1, new_data = future_months1) %>%
  bind_cols(future_months1)

# Combine actual and forecasted for plotting
combined_months1 <- bind_rows(
  monthly_counts_combined %>% select(month, clinical_count) %>% mutate(source = "Actual"),
  future_predictions1 %>% rename(clinical_count = .pred) %>% mutate(source = "Forecast")
)

# Plot actual vs forecast
seasonalpreds <- ggplot(combined_months1, aes(x = month, y = clinical_count, color = source)) +
  geom_line() +
  labs(title = "Actual and Forecasted Clinical Isolate Counts",
       y = "Clinical Count", x = "Month") +
  scale_color_manual(values = c("Actual" = "black", "Forecast" = "blue")) +
  theme_minimal()

print(seasonalpreds)

# Extract the fitted glm model from the workflow
poisson_fit1_glm <- extract_fit_engine(poisson_fit1)

# Calculate deviance residuals
deviance_residuals_clinical <- residuals(poisson_fit1_glm, type = "deviance")

# Calculate Pearson residuals
pearson_residuals_clinical <- residuals(poisson_fit1_glm, type = "pearson")

# Plot deviance residuals

residuals_df <- train_data2 %>%
  mutate(
    deviance_residual = deviance_residuals_clinical,
    pearson_residual = pearson_residuals_clinical,
    fitted = fitted(poisson_fit1_glm)
  )

# Deviance residual plot
ggplot(residuals_df, aes(x = fitted, y = deviance_residual)) +
  geom_point() +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  labs(title = "Deviance Residuals vs Fitted Values",
       x = "Fitted Values", y = "Deviance Residuals") +
  theme_minimal()

# Pearson residual plot
ggplot(residuals_df, aes(x = fitted, y = pearson_residual)) +
  geom_point() +
  geom_hline(yintercept = 0, color = "blue", linetype = "dashed") +
  labs(title = "Pearson Residuals vs Fitted Values",
       x = "Fitted Values", y = "Pearson Residuals") +
  theme_minimal()

# Estimate dispersion parameter
# Dispersion = sum of squared Pearson residuals / residual degrees of freedom
pearson_chisq <- sum(residuals_df$pearson_residual^2)
resid_df <- poisson_fit1_glm$df.residual
dispersion_estimate <- pearson_chisq / resid_df

print(paste("Estimated dispersion:", round(dispersion_estimate, 3)))

@@

#TEMPORAL X FOOD

```{r}
# Filling in gaps with 0 for food_count
monthly_counts_combined <- monthly_counts_combined %>%
  complete(month = seq.Date(min(month), max(month), by = "month"),
           fill = list(food_count = 0))

# Feature engineering: extracting the month from the data
monthly_counts_combined <- monthly_counts_combined %>%
  mutate(month_name = factor(month(month, label = TRUE), ordered = FALSE))

# Splitting the data
set.seed(123)
data_split2 <- initial_time_split(monthly_counts_combined, prop = 0.8)
train_data2 <- training(data_split2)
test_data2  <- testing(data_split2)

# Create a recipe for food_count
poisson_recipe_food <- recipe(food_count ~ month_name, data = train_data2) %>%
  step_dummy(all_nominal_predictors())

# Select Poisson model
poisson_model_food <- poisson_reg(
  mode = "regression",
  penalty = NULL,
  mixture = NULL,
  engine = "glm"
)

# Create a workflow
poisson_wf_food <- workflow() %>%
  add_model(poisson_model_food) %>%
  add_recipe(poisson_recipe_food)

# Fit the model
poisson_fit_food <- poisson_wf_food %>% fit(data = train_data2)

# Make predictions
predictions_food <- predict(poisson_fit_food, new_data = train_data2) %>%
  bind_cols(train_data2)

# Evaluate performance
library(yardstick)
metric_table2 <- metrics(predictions_food, truth = food_count, estimate = .pred)
save_table_location <- here::here("results","tables","metric_table2.rds")
saveRDS(metric_table2, file = save_table_location)

# Predict future food isolate counts
last_month <- max(monthly_counts_combined$month)

future_months_food <- tibble(
  month = seq.Date(from = last_month %m+% months(1), by = "month", length.out = 12)
) %>%
  mutate(month_name = factor(month(month, label = TRUE), ordered = FALSE))

# Future predictions
future_predictions_food <- predict(poisson_fit_food, new_data = future_months_food) %>%
  bind_cols(future_months_food)

# Combine actual and forecasted for plotting
combined_months_food <- bind_rows(
  monthly_counts_combined %>% select(month, food_count) %>% mutate(source = "Actual"),
  future_predictions_food %>% rename(food_count = .pred) %>% mutate(source = "Forecast")
)

# Plot actual vs forecast
pmodel_food <- ggplot(combined_months_food, aes(x = month, y = food_count, color = source)) +
  geom_line() +
  labs(title = "Actual and Forecasted Food Isolate Counts",
       y = "Food Count", x = "Month") +
  scale_color_manual(values = c("Actual" = "black", "Forecast" = "blue")) +
  theme_minimal()

# Print and save the plot
print(pmodel_food)
ggsave("/Users/alexisgonzalez/Desktop/MADA/Gonzalez-MADA-project/results/figures/pmodel_food.png",
       plot = pmodel_food, width = 8, height = 6, dpi = 300)

# Extract the fitted glm model
poisson_fit_food_glm <- extract_fit_engine(poisson_fit_food)

# Calculate deviance and Pearson residuals
deviance_residuals_food <- residuals(poisson_fit_food_glm, type = "deviance")
pearson_residuals_food <- residuals(poisson_fit_food_glm, type = "pearson")

# Create residuals data frame
residuals_df_food <- train_data2 %>%
  mutate(
    deviance_residual = deviance_residuals_food,
    pearson_residual = pearson_residuals_food,
    fitted = fitted(poisson_fit_food_glm)
  )

# Deviance residual plot
ggplot(residuals_df_food, aes(x = fitted, y = deviance_residual)) +
  geom_point() +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  labs(title = "Deviance Residuals vs Fitted Values (Food)",
       x = "Fitted Values", y = "Deviance Residuals") +
  theme_minimal()

# Pearson residual plot
ggplot(residuals_df_food, aes(x = fitted, y = pearson_residual)) +
  geom_point() +
  geom_hline(yintercept = 0, color = "blue", linetype = "dashed") +
  labs(title = "Pearson Residuals vs Fitted Values (Food)",
       x = "Fitted Values", y = "Pearson Residuals") +
  theme_minimal()

# Estimate dispersion
pearson_chisq_food <- sum(residuals_df_food$pearson_residual^2)
resid_df_food <- poisson_fit_food_glm$df.residual
dispersion_estimate_food <- pearson_chisq_food / resid_df_food

print(paste("Estimated dispersion (food model):", round(dispersion_estimate_food, 3)))

@@

#TEMPORAL X ENVIRONMENTAL

```{r}
# Filling in gaps with 0 for environmental_count
monthly_counts_combined <- monthly_counts_combined %>%
  complete(month = seq.Date(min(month), max(month), by = "month"),
           fill = list(environmental_count = 0))

# Feature engineering: extracting the month from the data
monthly_counts_combined <- monthly_counts_combined %>%
  mutate(month_name = factor(month(month, label = TRUE), ordered = FALSE))

# Splitting the data
set.seed(123)
data_split2 <- initial_time_split(monthly_counts_combined, prop = 0.8)
train_data2 <- training(data_split2)
test_data2  <- testing(data_split2)

# Create a recipe for environmental_count
poisson_recipe_env <- recipe(environmental_count ~ month_name, data = train_data2) %>%
  step_dummy(all_nominal_predictors())

# Select Poisson model
poisson_model_env <- poisson_reg(
  mode = "regression",
  penalty = NULL,
  mixture = NULL,
  engine = "glm"
)

# Create a workflow
poisson_wf_env <- workflow() %>%
  add_model(poisson_model_env) %>%
  add_recipe(poisson_recipe_env)

# Fit the model
poisson_fit_env <- poisson_wf_env %>% fit(data = train_data2)

# Make predictions
predictions_env <- predict(poisson_fit_env, new_data = train_data2) %>%
  bind_cols(train_data2)

# Evaluate performance
library(yardstick)
metric_table3 <- metrics(predictions_env, truth = environmental_count, estimate = .pred)
save_table_location <- here::here("results","tables","metric_table3.rds")
saveRDS(metric_table3, file = save_table_location)
# Predict future environmental isolate counts
last_month <- max(monthly_counts_combined$month)

future_months_env <- tibble(
  month = seq.Date(from = last_month %m+% months(1), by = "month", length.out = 12)
) %>%
  mutate(month_name = factor(month(month, label = TRUE), ordered = FALSE))

# Future predictions
future_predictions_env <- predict(poisson_fit_env, new_data = future_months_env) %>%
  bind_cols(future_months_env)

# Combine actual and forecasted for plotting
combined_months_env <- bind_rows(
  monthly_counts_combined %>% select(month, environmental_count) %>% mutate(source = "Actual"),
  future_predictions_env %>% rename(environmental_count = .pred) %>% mutate(source = "Forecast")
)

# Plot actual vs forecast
pmodel_env <- ggplot(combined_months_env, aes(x = month, y = environmental_count, color = source)) +
  geom_line() +
  labs(title = "Actual and Forecasted Environmental Isolate Counts",
       y = "Environmental Count", x = "Month") +
  scale_color_manual(values = c("Actual" = "black", "Forecast" = "blue")) +
  theme_minimal()

# Print and save the plot
print(pmodel_env)
ggsave("/Users/alexisgonzalez/Desktop/MADA/Gonzalez-MADA-project/results/figures/pmodel_env.png",
       plot = pmodel_env, width = 8, height = 6, dpi = 300)

#RESIDUALS AND DISPERSION

# Extract the fitted glm model
poisson_fit_env_glm <- extract_fit_engine(poisson_fit_env)

# Calculate deviance and Pearson residuals
deviance_residuals_env <- residuals(poisson_fit_env_glm, type = "deviance")
pearson_residuals_env <- residuals(poisson_fit_env_glm, type = "pearson")

# Create residuals data frame
residuals_df_env <- train_data2 %>%
  mutate(
    deviance_residual = deviance_residuals_env,
    pearson_residual = pearson_residuals_env,
    fitted = fitted(poisson_fit_env_glm)
  )

# Deviance residual plot
ggplot(residuals_df_env, aes(x = fitted, y = deviance_residual)) +
  geom_point() +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  labs(title = "Deviance Residuals vs Fitted Values (Environmental)",
       x = "Fitted Values", y = "Deviance Residuals") +
  theme_minimal()

# Pearson residual plot
ggplot(residuals_df_env, aes(x = fitted, y = pearson_residual)) +
  geom_point() +
  geom_hline(yintercept = 0, color = "blue", linetype = "dashed") +
  labs(title = "Pearson Residuals vs Fitted Values (Environmental)",
       x = "Fitted Values", y = "Pearson Residuals") +
  theme_minimal()

# Estimate dispersion
pearson_chisq_env <- sum(residuals_df_env$pearson_residual^2)
resid_df_env <- poisson_fit_env_glm$df.residual
dispersion_estimate_env <- pearson_chisq_env / resid_df_env

print(paste("Estimated dispersion (environmental model):", round(dispersion_estimate_env, 3)))

@@

#Negative Binomial

#CLINICAL
```{r}
# Prepare data
clinical_counts <- monthly_counts_combined %>%
  select(month, clinical_count) %>%
  complete(month = seq.Date(min(month), max(month), by = "month"),
           fill = list(clinical_count = 0)) %>%
  mutate(month_name = factor(month(month, label = TRUE), ordered = FALSE))

# Split
set.seed(123)
clinical_split <- initial_time_split(clinical_counts, prop = 0.8)
clinical_train <- training(clinical_split)
clinical_test  <- testing(clinical_split)

# Recipe
clinical_recipe <- recipe(clinical_count ~ month_name, data = clinical_train) %>%
  step_dummy(all_nominal_predictors())

# Model
clinical_nb_model <- linear_reg(mode = "regression") %>%
  set_engine("glm", family = MASS::negative.binomial(theta = 3.5))

# Workflow
clinical_nb_wf <- workflow() %>%
  add_model(clinical_nb_model) %>%
  add_recipe(clinical_recipe)

# Fit
clinical_nb_fit <- clinical_nb_wf %>%
  fit(data = clinical_train)

# Training predictions + metrics
clinical_preds <- predict(clinical_nb_fit, new_data = clinical_train) %>%
  bind_cols(clinical_train)

metric_table4<-metrics(clinical_preds, truth = clinical_count, estimate = .pred)
save_table_location <- here::here("results","tables","metric_table4.rds")
saveRDS(metric_table4, file = save_table_location)

##DISPERSION##
# Extract the fitted glm object
nb_fit_engine <- extract_fit_engine(clinical_nb_fit)

# Calculate Pearson residuals
pearson_resids <- residuals(nb_fit_engine, type = "pearson")

# Residual degrees of freedom
resid_df <- nb_fit_engine$df.residual

# Dispersion estimate
dispersion <- sum(pearson_resids^2) / resid_df

# Print
print(paste("Estimated Dispersion (Negative Binomial):", round(dispersion, 3)))

nmodel_clinical <- ggplot(clinical_preds, aes(x = month)) +
  geom_line(aes(y = clinical_count, color = "Actual")) +
  geom_line(aes(y = .pred, color = "Predicted")) +
  labs(
    title = "Actual and Predicted Clinical Isolate Counts (Negative Binomial Model)",
    x = "Month",
    y = "Clinical Count"
  ) +
  scale_color_manual(
    name = "Source",
    values = c("Actual" = "black", "Predicted" = "blue")
  ) +
  theme_minimal()

# Display the plot
nmodel_clinical

ggsave("/Users/alexisgonzalez/Desktop/MADA/Gonzalez-MADA-project/results/figures/nmodel_clinical.png",
       plot = nmodel_clinical, width = 8, height = 6, dpi = 300)
```

#FOOD

```{r}
# Prepare data
food_counts <- monthly_counts_combined %>%
  select(month, food_count) %>%
  complete(month = seq.Date(min(month), max(month), by = "month"),
           fill = list(food_count = 0)) %>%
  mutate(month_name = factor(month(month, label = TRUE), ordered = FALSE))

# Split
set.seed(123)
food_split <- initial_time_split(food_counts, prop = 0.8)
food_train <- training(food_split)
food_test  <- testing(food_split)

# Recipe
food_recipe <- recipe(food_count ~ month_name, data = food_train) %>%
  step_dummy(all_nominal_predictors())

# Model
food_nb_model <- linear_reg(mode = "regression") %>%
  set_engine("glm", family = MASS::negative.binomial(theta = 2))

# Workflow
food_nb_wf <- workflow() %>%
  add_model(food_nb_model) %>%
  add_recipe(food_recipe)

# Fit
food_nb_fit <- food_nb_wf %>%
  fit(data = food_train)

# Training predictions + metrics
food_preds <- predict(food_nb_fit, new_data = food_train) %>%
  bind_cols(food_train)

metrics_table5 <- metrics(food_preds, truth = food_count, estimate = .pred)
save_table_location <- here::here("results","tables","metrics_table5.rds")
saveRDS(metrics_table5, file = save_table_location)
# Extract the fitted glm object from the workflow
food_nb_engine <- extract_fit_engine(food_nb_fit)

# Calculate Pearson residuals
pearson_resids_food <- residuals(food_nb_engine, type = "pearson")

# Residual degrees of freedom
resid_df_food <- food_nb_engine$df.residual

# Dispersion estimate
dispersion_food <- sum(pearson_resids_food^2) / resid_df_food

# Print the dispersion
print(paste("Estimated Dispersion (Negative Binomial - Food Model):", round(dispersion_food, 3)))

nmodel_food <- ggplot(food_preds, aes(x = month)) +
  geom_line(aes(y = food_count, color = "Actual")) +
  geom_line(aes(y = .pred, color = "Predicted")) +
  labs(
    title = "Actual and Predicted Food Isolate Counts (Negative Binomial Model)",
    x = "Month",
    y = "Food Count"
  ) +
  scale_color_manual(
    name = "Source",
    values = c("Actual" = "black", "Predicted" = "blue")
  ) +
  theme_minimal()

# Print the plot
nmodel_food

ggsave("/Users/alexisgonzalez/Desktop/MADA/Gonzalez-MADA-project/results/figures/nmodel_food.png",
       plot = nmodel_food, width = 8, height = 6, dpi = 300)

```
#ENVIRONMENTAL
```{r}

# Prepare data
environmental_counts <- monthly_counts_combined %>%
  select(month, environmental_count) %>%
  complete(month = seq.Date(min(month), max(month), by = "month"),
           fill = list(environmental_count = 0)) %>%
  mutate(month_name = factor(month(month, label = TRUE), ordered = FALSE))

# Split
set.seed(123)
env_split <- initial_time_split(environmental_counts, prop = 0.8)
env_train <- training(env_split)
env_test  <- testing(env_split)

# Recipe
env_recipe <- recipe(environmental_count ~ month_name, data = env_train) %>%
  step_dummy(all_nominal_predictors())

# Model
env_nb_model <- linear_reg(mode = "regression") %>%
  set_engine("glm", family = MASS::negative.binomial(theta = 1))

# Workflow
env_nb_wf <- workflow() %>%
  add_model(env_nb_model) %>%
  add_recipe(env_recipe)

# Fit
env_nb_fit <- env_nb_wf %>%
  fit(data = env_train)

# Training predictions + metrics
env_preds <- predict(env_nb_fit, new_data = env_train) %>%
  bind_cols(env_train)

metrics_table6 <- metrics(env_preds, truth = environmental_count, estimate = .pred)
save_table_location <- here::here("results","tables","metrics_table6.rds")
saveRDS(metrics_table6, file = save_table_location)
# Extract the fitted glm object from the workflow
env_nb_engine <- extract_fit_engine(env_nb_fit)

# Calculate Pearson residuals
pearson_resids_env <- residuals(env_nb_engine, type = "pearson")

# Residual degrees of freedom
resid_df_env <- env_nb_engine$df.residual

# Dispersion estimate
dispersion_env <- sum(pearson_resids_env^2) / resid_df_env

# Print the dispersion
print(paste("Estimated Dispersion (Negative Binomial - Environmental Model):", round(dispersion_env, 3)))

nmodel_env <- ggplot(env_preds, aes(x = month)) +
  geom_line(aes(y = environmental_count, color = "Actual")) +
  geom_line(aes(y = .pred, color = "Predicted")) +
  labs(
    title = "Actual and Predicted Environmental Isolate Counts (Negative Binomial Model)",
    x = "Month",
    y = "Environmental Count"
  ) +
  scale_color_manual(
    name = "Source",
    values = c("Actual" = "black", "Predicted" = "blue")
  ) +
  theme_minimal()

# Display the plot
nmodel_env

ggsave("/Users/alexisgonzalez/Desktop/MADA/Gonzalez-MADA-project/results/figures/nmodel_env.png",
       plot = nmodel_env, width = 8, height = 6, dpi = 300)

@@
